---
title: "Activity Monitoring Analysis"
author: "Shouryaman Singh"
date: "2026-04-18"
output:
  html_document: default
  md_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

```{r}
data <- read.csv("activity.csv")
head(data)
```

---

## What is mean total number of steps taken per day?

```{r}
steps_per_day <- data %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = TRUE))
```

```{r}
hist(steps_per_day$total_steps,
     main = "Total Steps Per Day",
     xlab = "Steps",
     col = "blue")
```

```{r}
mean_steps <- mean(steps_per_day$total_steps)
median_steps <- median(steps_per_day$total_steps)

mean_steps
median_steps
```

The mean number of steps per day is **`r round(mean_steps,2)`** and the median is **`r round(median_steps,2)`**.

---

## What is the average daily activity pattern?

```{r}
avg_interval <- data %>%
  group_by(interval) %>%
  summarise(avg_steps = mean(steps, na.rm = TRUE))
```

```{r}
plot(avg_interval$interval, avg_interval$avg_steps,
     type = "l",
     xlab = "Interval",
     ylab = "Average Steps")
```

```{r}
avg_interval[which.max(avg_interval$avg_steps), ]
```

---

## Imputing missing values

```{r}
sum(is.na(data$steps))
```

```{r}
data_filled <- data

for(i in 1:nrow(data_filled)){
  if(is.na(data_filled$steps)){
    data_filled$steps[i] <- avg_interval$avg_steps[
      avg_interval$interval == data_filled$interval[i]
    ]
  }
}
```

```{r}
steps_per_day2 <- data_filled %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))
```

```{r}
hist(steps_per_day2$total_steps,
     main = "Total Steps Per Day (After Imputation)",
     xlab = "Steps",
     col = "green")
```

```{r}
mean2 <- mean(steps_per_day2$total_steps)
median2 <- median(steps_per_day2$total_steps)

mean2
median2
```

After imputing missing values, the mean is **`r round(mean2,2)`** and the median is **`r round(median2,2)`**.

---

## Are there differences in activity patterns between weekdays and weekends?

```{r}
data_filled$date <- as.Date(data_filled$date)

data_filled$day_type <- ifelse(
  weekdays(data_filled$date) %in% c("Saturday", "Sunday"),
  "Weekend",
  "Weekday"
)
```

```{r}
avg_daytype <- data_filled %>%
  group_by(interval, day_type) %>%
  summarise(avg_steps = mean(steps))
```

```{r}
ggplot(avg_daytype, aes(x = interval, y = avg_steps)) +
  geom_line() +
  facet_wrap(~day_type) +
  labs(title = "Activity Pattern: Weekday vs Weekend",
       x = "Interval",
       y = "Average Steps")
```

The activity patterns differ between weekdays and weekends.
